The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-02-20 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2473 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bcdd074e435d195cd5e1ea3b6c88485e988b70ba

%\newpage
\subsection{Tree}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have
been evaluated on data sets based on the tree structure with two different split variables 
($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
Figure~\ref{fig:dgp_tree}.

Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE\_scores performs
a bit worse but it still achieves an ARI above $0.5$ for the highest considered effect size
of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
slope parameter as it only considers residuals in its testing strategy.
For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
methods and both GUIDE versions fail to find the correct tree structure.
Based on these results it can be concluded once again that including all available scores in the
testing strategy leads to a better performance while binarizing scores and categorizing
split variables both do not cause any improvement but in many situations even have a negative
effect on the power of a testing strategies.

@@

To elaborate the impact of the choice of a pruning strategy
the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
applying postpruning (building very large trees without any significance level and employing 
cost-complexity pruning as in the CART algorithm). 
In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with prepruning,
again based on the Adjusted Rand Index.

@@

Linear models as presented in this study have been selected as they allow for a simple illustration
of the investigated subjects. However, the obtained results also hold for a variety of other models
as the introduced building blocks can be applied in the same way for all model types.
One specific feature of linear models is the orthogonality of the conditional expectation
$\mathbb{E}(Y|X)$ and the regressor $X$ which affects the capability of tests using only residuals
to capture shifts in the parameters. In particular, \cite{Ploberger+Kramer:1992} have shown that
residual-based tests are insensitive to parameter changes orthogonal to the mean regressor.
For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
Testing procedures purely based on residuals perform best in situations with shifts only in the
intercept, however, the costs in terms of power that come with including all scores are 
negligibly small. 
Furthermore, with only two parameters determining a linear model the importance of including scores 
regarding each parameter could already be pointed out clearly while for an increasing number of 
parameters in a model this aspect becomes even more important.

In general residual-based and score-based procedures are expected to perform equally well if all
parameters are highly correlated. On the other hand, if parameters are orthogonal changes that are 
not connected to the mean might be missed when only considering residuals. This can be avoided by
including scores which are a standard measurement of goodness of fit that can be easily employed.
Even though it also depends on the type of the applied residuals 
(see, e.g., \cite{Loh+Zheng:2013}, \cite{Chaudhuri+Loh:2002} 
and \cite{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made.

>>>>

%\newpage
\subsection{Tree}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have
been evaluated on data sets based on the tree structure with two different split variables 
($Z_1$ and $Z_2$) both with the same split point $\xi$ as illustrated in 
Figure~\ref{fig:dgp_tree}.

Figure~\ref{fig:tree_ari} shows that based on the resulting ARI values CTree and MOB perform 
almost equally well for the true split points $0$, $0.2$ and $0.5$. GUIDE\_scores performs
slightly worse but it still achieves an ARI above $0.5$ for the highest considered effect size
of $\delta = 1$. However, the original version of GUIDE is clearly behind which is mainly
due to the fact that it doesn't detect the first split in $Z_2$ corresponding to a changing
slope parameter as it only considers residuals in its testing strategy.
For a split point farther away from the center such as for $\xi = 0.8$ MOB outperforms all other
methods and both GUIDE versions fail to find the correct tree structure.
Based on these results it can be concluded once again that including all available scores in the
testing strategy leads to a better performance while binarizing scores and categorizing
split variables both do not cause any improvement but in many situations even have a negative
effect on the power of a testing strategy.

@@

To elaborate the impact of the choice of a pruning strategy
the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
applying postpruning (building very large trees without any significance level and employing 
cost-complexity pruning as in the CART algorithm \citep{Breiman+Friedman+Stone:1984}). 
In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with prepruning,
again based on the Adjusted Rand Index.

@@

Linear models as presented in this study have been selected as they allow for a simple illustration
of the investigated subjects. However, the obtained results also hold for a variety of other models
as the introduced building blocks can be applied in the same way for all model types.
Specific characteristics of linear models regarding structural changes have been investigated
in detail in previous literature.
\cite{Ploberger+Kramer:1992} have shown that residual-based tests can detect a change in the parameters
only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
the parameter vector $\bm{\beta}$ is shifted 
%to $\bm{\beta} + \bm{\Delta}$ 
by $\bm{\Delta} = (0, \Delta_1)^{\top}$.
For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
While this might be an extreme example similar scenarios can appear in many applications, see, e.g., 
Figure~2 in \cite{Loh+He+Man:2015} where an additional separation of the data in each node is
required in order to fit a reasonable model. This can be avoided by including model scores
allowing for the detection of changes not only in the location parameter but also in all other 
parameters, i.e., not only in the expected value but also in the variance.
In situations where shifts appear only in the intercept testing procedures purely based on residuals 
perform best. However, the cost in terms of power that comes with including all scores, in particular
with testing more than one parameter, is negligibly small.

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \cite{Loh+Zheng:2013}, \cite{Chaudhuri+Loh:2002} 
and \cite{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kramer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of of testing all parameters.

%Furthermore, with only two parameters determining a linear model the importance of including scores 
%regarding each parameter could already be pointed out clearly while for an increasing number of 
%parameters in a model this aspect becomes even more important.
